---
title: "VI"
author: "Yichen Ji"
date: "18/11/2021"
output: html_document
---
```{r}
rm(list = ls())
library(mvtnorm)  # for multivariate normal density
library(pracma)   # for 2d-integral, sqrtm
library(ggplot2)  # for plotting
library(tidyverse)
```

Logistic likelihood and Gaussian priors

Reference to the paper 'Conditionally conjugate mean–field variational Bayes for logistic models'
by Daniele Durante and Tommaso Rigon

MCMC as benchmark to compare

more covariates e.g. 5?

number of MCMC samples per replication = 5000
number of replications = 200

compare MSE of different methods

```{r}
# true parameter values i.e. betas
beta = c(1,1,1,1,1)

# prior hyperparameters
prior = list(mu = rep(0,5), Sigma = diag(1,5)) # zero mean and identity covariance matrix

# Variational Inference(SVI) parameter settings
iter = 5000 # number of iterations
tau = 1 # delay 
kappa = 0.75 # forgetting rate
```

```{r}
set.seed(123)

n = 20 # sample size
x1 = runif(n, -1, 1)
x2 = runif(n, -1, 1)
x3 = runif(n, -1, 1)
x4 = runif(n, -1, 1)
X = cbind(1, x1,x2,x3,x4) # the design matrix including the intercept
y = rbinom(n, 1, prob = plogis(X %*% beta))
```

use the quardratic lower bound proposed from Jaakkola and Jordan (2000)

```{r}
logit_CAVI = function(X,y,prior,tol = 1e-16, maxiter=10000){
  if (is.null(n = nrow(X))) stop("'X' should be a matrix, not a vector")
  
  # compute the log-determinant of a matrix
  ldet = function(X){
    if(!is.matrix(X)) return(log(X))
    determinant(X, logarithm = T)$modulus
  }
  
  lowerbound = numeric(maxiter) # ELBO
  p = ncol(X) # number of betas
  
  P = solve(prior$Sigma)
  mu = prior$mu
  Pmu = c(P %*% mu)
  Pdet = ldet(P)
  
  # initialization for omega = 0.25
  P_vb = crossprod(X*rep(0.25, n), X) + P
  Sigma_vb = solve(P_vb)
  mu_vb = Sigma_vb %*% (crossprod(X, y - 0.5) + Pmu)
  eta = c(X %*% mu_vb)
  xi = sqrt(eta^2 + rowSums(X %*% Sigma_vb * X))
  omega = tanh(xi/2) / (2 * xi)
  omega[is.nan(omega)] = 0.25
  
  lowerbound[1] = 0.5*p + 0.5*ldet(Sigma_vb) + 0.5*Pdet -
    0.5*t(mu_vb - mu)%*%P%*%(mu_vb - mu) +
    sum((y-0.5)*eta +log(plogis(xi)) - 0.5*xi) -
    0.5*sum(diag(P %*% Sigma_vb))
  
  # iterative procedure
  for (t in 2:maxiter){
    P_vb = crossprod(X*omega, X) + P
    Sigma_vb = solve(P_vb)
    mu_vb = Sigma_vb %*% (crossprod(X, y - 0.5) + Pmu)
    
    # update xi
    eta = c(X %*% mu_vb)
    xi = sqrt(eta^2 + rowSums(X %*% Sigma_vb * X))
    omega = tanh(xi/2) / (2 * xi)
    omega[is.nan(omega)] = 0.25
    
    lowerbound[t] = 0.5*p + 0.5*ldet(Sigma_vb) + 0.5*Pdet 
    - 0.5*t(mu_vb - mu)%*%P%*%(mu_vb - mu) + sum((y-0.5)*eta +
                                                   log(plogis(xi)) - 0.5*xi) - 
      0.5*sum(diag(P %*% Sigma_vb))
    
    if(abs(lowerbound[t] - lowerbound[t-1]) < tol) 
      return(list(mu = matrix(mu_vb,p,1),
                  Sigma = matrix(Sigma_vb,p,p),
                  Convergence = cbind(Iteration=(1:t)-1,
                                      Lowerbound=lowerbound[1:t]), 
                  xi=xi))
  }
  stop("The algorithm has not reached convergence")
}
```

```{r}
set.seed(123)

CAVI_output = logit_CAVI(X = X, y = y, prior = prior)

# Posterior distribution of the intercept with CAVI
beta0_CAVI = rnorm(iter, CAVI_output$mu[1], sqrt(CAVI_output$Sigma[1,1])) 

# Posterior distribution of the slope with CAVI
beta1_CAVI = rnorm(iter, CAVI_output$mu[2], sqrt(CAVI_output$Sigma[2,2])) 

beta2_CAVI = rnorm(iter, CAVI_output$mu[3], sqrt(CAVI_output$Sigma[3,3]))

beta3_CAVI = rnorm(iter, CAVI_output$mu[4], sqrt(CAVI_output$Sigma[4,4]))

beta4_CAVI = rnorm(iter, CAVI_output$mu[5], sqrt(CAVI_output$Sigma[5,5]))

# Posterior distribution of the intercept with SVI
# beta0_SVI  <- rnorm(10^4, SVI_output$mu[1], sqrt(SVI_output$Sigma[1,1]))   

# Posterior distribution of the slope with SVI
# beta1_SVI  <- rnorm(10^4, SVI_output$mu[2], sqrt(SVI_output$Sigma[2,2]))   

data_plot = data.frame(Posterior = c(beta0_CAVI,beta1_CAVI,beta2_CAVI,beta3_CAVI,beta4_CAVI), 
                       beta = rep(c("beta0","beta1",'beta2','beta3','beta4'),each=iter), 
                       Algorithm = rep(c("CAVI"),each=5*iter), 
                       Sample_size = n)
```

```{r}
round = 200
sample_mean = data.frame(beta0=double(),beta1=double(),beta2=double(),beta3=double(),beta4=double())
sample_var = data.frame(beta0=double(),beta1=double(),beta2=double(),beta3=double(),beta4=double())

i=1
while(i <= round){
  CAVI_output = logit_CAVI(X = X, y = y, prior = prior)
  
  beta0 = rnorm(iter, CAVI_output$mu[1], sqrt(CAVI_output$Sigma[1,1]))
  mean_beta0 = mean(beta0)
  var_beta0 = var(beta0)
  
  beta1 = rnorm(iter, CAVI_output$mu[2], sqrt(CAVI_output$Sigma[2,2]))
  mean_beta1 = mean(beta1)
  var_beta1 = var(beta1)
  
  beta2 = rnorm(iter, CAVI_output$mu[3], sqrt(CAVI_output$Sigma[3,3]))
  mean_beta2 = mean(beta2)
  var_beta2 = var(beta2)
  
  beta3 = rnorm(iter, CAVI_output$mu[4], sqrt(CAVI_output$Sigma[4,4]))
  mean_beta3 = mean(beta3)
  var_beta3 = var(beta3)
  
  beta4 = rnorm(iter, CAVI_output$mu[5], sqrt(CAVI_output$Sigma[5,5]))
  mean_beta4 = mean(beta4)
  var_beta4 = var(beta4)
  
  sample_mean = add_row(sample_mean, beta0 = mean_beta0,
                          beta1 = mean_beta1,
                          beta2 = mean_beta2,
                          beta3 = mean_beta3,
                          beta4 = mean_beta4)
  sample_var = add_row(sample_var, beta0 = var_beta0,
                          beta1 = var_beta1,
                          beta2 = var_beta2,
                          beta3 = var_beta3,
                          beta4 = var_beta4)
  i = i+1
}

# mean-squared error of each round for each beta
MSE = (sample_mean - beta)^2 + sample_var
round_mean_MSE = sapply(MSE, mean)
round_mean_MSE
```


```{r}
ggplot(data=data_plot, 
       aes(x = as.factor(Sample_size), 
           y = Posterior, fill=Algorithm)) +
  facet_grid(~beta) + 
  geom_boxplot(alpha=0.7) + 
  theme_bw() + 
  scale_fill_grey() + 
  geom_hline(yintercept=1, linetype="dotted") + 
  xlab("Sample size") + 
  ylab("Logistic Regression Coefficient")

```
```{r}
nn <- c(1000, 5000, 10000) # setting the sample size
data_plots = data.frame()
for(n in nn){
  set.seed(123)      # Set the seed to make this experiment reproducible
  x1 <- runif(n,-1,1)
  x2 <- runif(n,-1,1)
  x3 <- runif(n,-1,1)
  x4 <- runif(n,-1,1)
  X <- cbind(1,x1,x2,x3,x4)    
  y <- rbinom(n,1,prob = plogis(X%*%beta))

  set.seed(1010)     # Set the seed to make this experiment reproducible
  CAVI_output <- logit_CAVI(X = X, y = y, prior = prior) # CAVI 
  
  #SVI_output  <- logit_SVI(X = X, y = y,  prior = prior,  
                           #iter = iter, tau = tau, kappa = kappa) # SVI

  set.seed(100)
  
  # Posterior distribution of the intercept with CAVI
  beta0_CAVI <- rnorm(10^4, CAVI_output$mu[1], sqrt(CAVI_output$Sigma[1,1]))
  
  # Posterior distribution of the slope with CAVI
  beta1_CAVI <- rnorm(10^4, CAVI_output$mu[2], sqrt(CAVI_output$Sigma[2,2])) 
  
  beta2_CAVI = rnorm(10^4, CAVI_output$mu[3], sqrt(CAVI_output$Sigma[3,3]))

  beta3_CAVI = rnorm(10^4, CAVI_output$mu[4], sqrt(CAVI_output$Sigma[4,4]))

  beta4_CAVI = rnorm(10^4, CAVI_output$mu[5], sqrt(CAVI_output$Sigma[5,5]))
  
  # Posterior distribution of the intercept with SVI
  #beta0_SVI  <- rnorm(10^4, SVI_output$mu[1], sqrt(SVI_output$Sigma[1,1]))  
  
  # Posterior distribution of the slope with SVI
  #beta1_SVI  <- rnorm(10^4, SVI_output$mu[2], sqrt(SVI_output$Sigma[2,2]))   

  data_plots <- rbind(data_plots,data.frame(Posterior = c(beta0_CAVI,beta1_CAVI,
                                                         beta2_CAVI, beta3_CAVI, beta4_CAVI),
                                          beta = rep(rep(c("beta0","beta1",'beta2','beta3','beta4'),
                                                         each=10^4),5), 
                                          Algorithm = rep(c("CAVI"),
                                                          each=5*10^4), 
                                          Sample_size = n))
}
```

```{r}
ggplot(data=data_plots, 
       aes(x = Posterior, color = beta, fill=beta)) + 
  geom_histogram(alpha=0.5,binwidth = 0.005)+
  facet_wrap(~Sample_size,ncol = 1)+
  theme_minimal()+
  geom_vline(aes(xintercept = 1))
```
```{r}
# color divide by sample size for each beta
ggplot(data=data_plots, 
       aes(x = Posterior, color = Sample_size)) + 
  geom_histogram(alpha=0.5,binwidth = 0.005)+
  facet_wrap(~beta,ncol = 1)+
  theme_minimal()+
  geom_vline(aes(xintercept = 1))
```


```{r}
ggplot(data=data_plots, 
       aes(x = as.factor(Sample_size), 
           y = Posterior, fill=Algorithm)) + 
  facet_grid(~beta) + 
  geom_boxplot(alpha=0.7) + 
  theme_bw() + 
  scale_fill_grey() + 
  geom_hline(yintercept=1, linetype="dotted") + 
  xlab("Sample size") + 
  ylab("Regression Coefficient")

```


```{r}
### KL Divergence
KL = function(p,q,low = -Inf, up = Inf){
  f = function(x) p(x)*(p(x,T)-q(x,T)) # T means taking logarithm
  integrate(f, lower = low, upper = up)
}


# KL between two normal distributions
curve(dnorm(x), from = -9, to = 11) # mean = 0
curve(dnorm(x,2), from = -9, to = 11, add = T, col = 2) # mean = 2
p = function(x, lg=F) dnorm(x, log=lg)
q = function(x, lg=F) dnorm(x, 2, log=lg)

KL(p,q)
KL(q,p)
# in this case KL(p,q) = KL(q,p), but not in general


# KL between normal and t distribution
curve(dnorm(x), from = -5, to = 5)
curve(dt(x,1), from = -5, to = 5, add = T, col=2)
p = function(x, lg=F) dnorm(x, log=lg)
q = function(x, lg=F) dt(x,1,log=lg)

KL(p,q)
```


```{r}

### Minimize KL-divergence between a correlated 2D Gaussian

# 2D KL Divergence
KL_2D = function(q,p){
  f = function(x,y) q(x,y) * (q(x,y,T) - p(x,y,T))
  integral2(f, -20, 20, -20, 20)$Q
}


# covariance matrix of the true posterior
Sigma = matrix(c(1, 0.9, 0.9, 1), ncol = 2)

# true posterior
posterior = function(x, y, lg=F){
  dmvnorm(matrix(c(x,y), ncol = 2), mean = rep(0,2), sigma = Sigma, log = lg)
}

# Sanity checks
posterior(0,0)
integral2(posterior, -10, 10, -10, 10)$Q
```

```{r}
# Mean Field Variational Family i.e. product of approximate distributions
# here the function form is still Normal

# variance of two Gaussians as hyper-parameter
s1 = 2
s2 = 2


qq = function(x,y,lg=F){
  # independence assumption of mean-field
  if(lg){
    dnorm(x, 0, s1, log=lg) + dnorm(y, 0, s2, log=lg)
  } else{
    dnorm(x,0,s1)* dnorm(y,0,s2) 
  }
}

# sanity check
qq(0,0)
integral2(qq, -10, 10, -10, 10)$Q

# sum up the above into one function

func = function(t){ # t contains separate variational parameters
  qt = function(x,y,lg=F){
    if(lg){
      dnorm(x,0,t[1], log=T) + dnorm(y,0,t[2],log=T)
    } else{
      dnorm(x,0,t[1]) * dnorm(y,0,t[2])
    }
  }
  KL_2D(qt,posterior)
}

func(c(0.4,0.9))
```

```{r}
KL_2D(qq,posterior)
KL_2D(posterior, qq)
```

```{r}
# Plot the two distributions
circleFun = function(center=c(0,0), diameter=1, npoints=100){
  r = diameter / 2
  tt = seq(0,2*pi,length.out = npoints)
  xx = center[1] + r * cos(tt)
  yy = center[2] + r * sin(tt)
  return(data.frame(x = xx, y = yy))
}

plotCircle = function(s) {
  dat = circleFun(c(0,0),4,npoints = 100) 
  # represent the circle of 2*sigma of MVN(c(0,0), diag(1,1))
  
  dat1 = as.matrix(dat) %*% sqrtm(Sigma)$B  
  # represents the posterior MVN(c(0,0), Sigma)
  
  dat2 = as.matrix(dat) %*% diag(s)
  # represents the variational family MVN(c(0,0), diag(s1, s2))
  
  ggplot(as.data.frame(dat1),aes(x=V1,y=V2)) + theme_light() + 
    ylim(-5,5) + xlim(-5,5) + 
    xlab("x") +ylab("y") + 
    geom_polygon(color="#2E9FDF", 
                 fill="#2E9FDF", 
                 alpha=0.3) +
    geom_polygon(data=as.data.frame(dat2), 
                 aes(x=V1,y=V2, color="#ff7d9d"), 
                 fill="#ff7d9d", 
                 alpha=0.3, 
                 show.legend = F)
}

# Starting value - s1 = 1, s2 = 1
plotCircle(c(1,1))

# Minimize KL(q,post)
fit = optim(par=c(1,1), fn=func, method="L-BFGS-B", lower=0)
fit$par

plotCircle(fit$par)
```

```{r}
# know the posterior up to some normalizing constant
NC = 0.5
post.unk = function(x,y,lg=F) {
  if (lg) {
    log(NC) + posterior(x,y,T)
  } else {
    NC*posterior(x,y)
  }
}
integral2(post.unk, -10, 10, -10, 10)$Q
post.unk(0,0)
KL_2D(qq, post.unk)

# Wrap the q function so we can give it different values of sigma's and optimize it
func2 = function(t) {
  qt = function(x,y,lg=F) {
    if (lg) {
      dnorm(x,0,t[1],log=T) + dnorm(y,0,t[2],log=T)
    } else {
      dnorm(x,0,t[1])*dnorm(y,0,t[2])
    }
  }
  KL_2D(qt, post.unk) # note that since we are giving it only the joint (unnormalized posterior)
                   # this is equivalent to -ELBO
}

# optim finds the minimum, so using this finds the maximum of the ELBO
fit = optim(par=c(1,1), fn=func2, method="L-BFGS-B", lower=0)
fit$par

plotCircle(fit$par)
```


```{r}
# Variational Inference 
# We want to minimize the KL divergence between two distributions
# By using Gradient Descent/Ascent

# Example: P(x) - Real ~ N(0,1)
x = rnorm(100) # observed values
# Q(x) - approximation ~ N(mu, sigma^2)
# start with randomized values
mu.0 = 5
sigma.sq.0 = 3
# KL(P||Q) = E[log P(x)] - E[log Q(x)]  (E w.r.t. P) we want to minimize this
# 1st term doesn't depend on parameters so we want to maximize 2nd term
# Gradient (w.r.t. mu and sigma^2) comes into the expectations
# Approximate E with sample average
# ( KL(Q||P) wouldn't work, as we don't have any way to approximate P(x) )
logQ = function(x, mu, sigma.sq) {
  -.5*log(2*pi)-.5*log(sigma.sq)-.5*(x-mu)^2/sigma.sq
}
dlogQ.dMu = function(x, mu, sigma.sq) {
  (x-mu)/sigma.sq
}
dlogQ.dS = function(x, mu, sigma.sq) {
  .5*((x-mu)/sigma.sq)^2-1/(2*sigma.sq)
}

# Gradient ascent
ls = 0.5 # learning step
mu.t = mu.0
sigma.sq.t = sigma.sq.0
tol = 1e-4
step.mu = step.sigma = 1
while((abs(step.mu) > tol) || (abs(step.sigma) > tol)) {
  step.mu = ls*mean(dlogQ.dMu(x, mu.t, sigma.sq.t))
  mu.t = mu.t + step.mu
  step.sigma = ls*mean(dlogQ.dS(x, mu.t, sigma.sq.t))
  sigma.sq.t = sigma.sq.t + step.sigma
}

curve(dnorm(x,mu.t,sqrt(sigma.sq.t)), from=-3, to=3)
curve(dnorm(x),from=-3, to=3, add=T, col=2)

# results are equal to the Maximum-Likelihood estimators
n = length(x)
c(mean(x), mu.t)
c(var(x)*(n-1)/n, sigma.sq.t)
```










